---
title: 'Pristina TranscriptomicLandscape 02: TF expression analysis'
author: "Alberto Perez-Posada @apposada"
date: "30/11/2022"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}

dir <- '/mnt/sda/alberto/colabos/pristina/pristina_transcriptomic_landscape/'
fcha <- function(){ gsub("-","",Sys.Date()) }

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = dir)
options(scipen=999)

```

## Loading Necessary Packages

```{r warning = FALSE, message=FALSE}
library(data.table)
library(reshape2)
library(dplyr)
library(tidyverse)
library(ComplexHeatmap)
library(ggplot2)
library(circlize)
library(RColorBrewer)
library(viridis)
library(colorspace)

```


## Loading necessary code (CHANGE PATHS)

```{r, warning = FALSE, message=FALSE}
#courtesy from @Rich Scriven https://stackoverflow.com/questions/26045478/source-r-scripts-in-different-folders
sourceFolder <- function(folder, recursive = FALSE, ...) 
{ 
    files <- list.files(folder, pattern = "[.][rR]$", 
                        full.names = TRUE, recursive = recursive)
    if (!length(files))
        stop(simpleError(sprintf('No R files in folder "%s"', folder)))
    src <- invisible(lapply(files, source, ...))
    message(sprintf('%s files sourced from folder "%s"', length(src), folder))
}

sourceFolder(
  "/mnt/sda/alberto/projects/dev/comparABle/code/functions/", # change for definitive path once its final
  recursive = TRUE
  )
# source("/mnt/sda/alberto/projects/dev/minitools_R/code/functions/")

source('./code/R_functions/TF_analysis_functions/barplotcluster.R') # change for definitive path once its final
```


## Data Load

We start with our previous session.

```{r load_data}
load(
  "./outputs/rda/01_plei_dataload.rda"
)
```

We will also load the TF annotation of Pristina leidyi.

```{r load_TF_annotation}
plei_tfs <- read.table(
  "./data/03_TFdata/20221025_pristina_TFs_curated.tsv",
  header = T)[,1:2] #load column of plei ID and TF class
head(plei_tfs)
```

We subset the gene expression pseudo-bulk matrix to retrieve expression from the TFs.

```{r}
plei_tfs_cpm <-
  plei_cpm[
    rownames(plei_cpm) %in% plei_tfs$id,
    1:49
    ]

colnames(plei_tfs_cpm) <- colnames(plei_cpm)[1:49]

plei_tfs_cpm <- as.data.frame(plei_tfs_cpm)

plei_tfs_cpm_hvg <- 
  plei_tfs_cpm[
    rownames(plei_tfs_cpm) %in% plei_hvg$id,
    ]
```

We can browse the expression level of different TFs using this function.

```{r}
plot_tf_plei <- function(x){
  if(x %in% rownames(plei_tfs_cpm)) {
    barplot(
      height=unlist(c(
        plei_tfs_cpm[
          grep(
            paste("^",x,"$",sep=""),
            rownames(plei_tfs_cpm),
          ),
          1:49
        ]
      )),
      col = plei_ctypes_col$col,
      border = "#2F2F2F",
      las=2,
      cex.names=0.7,
      main= paste(
        x,
        " (",
        plei_tfs[grep(x,plei_tfs$id),2],
        ")\n",
        sep=""
      ),
      ylab="counts per million per cluster"
    )} else {
      stop("Name not in list of TFs.")
    }
}

```

For example, the hnf4 gene;

```{r}
hnf4 <- "PrileiEVm006891t1"
plot_tf_plei(hnf4)

# -gut,hnf4 : 006891, 004665, 008019
# -muscle: myoD: 008071, 014648
# -neurons: myt1l: 000431, pou6f :003917
# -UMAP feature of novel cell types:
#   -lipoxygenase+ 001926
# -globin+ 002870 005469
# -eleocytes 005230 – ets4
# -vigilin+ 005896, 006907
# -polycystin 010161

```

As we have expression data of many transcription factors, we can visualise the global
patterns of expression using heatmaps.
We will do so by scaling the log-transformed expression of TFs to obtain a z-score.

```{r}
plei_tfs_zsco <- 
  t(
    scale(
      t(
        log(plei_tfs_cpm[,1:49]+1) # we add 1 for visualisation only
        )
      )
    )
```

And using the ComplexHeatmap package:

```{r echo=FALSE}
# Color Palettes

#Color palette for heatmap of TF-TF Spearman correlation
plei_tfs_heat_col <- colorRamp2(
  c( # breaks, clipped
    seq(-0.4,0.2,len=10),
    seq(0.3,0.5,len=10)
  ),
  colorRampPalette( # colors
    rev(brewer.pal(7,"RdYlBu"))
  )(20)
)

# Color palette for heatmap of expression
col_plei_expr_zsco_hm <- colorRamp2(
  c(1:6,6.5), # breaks, clipped
  rev(sequential_hcl(7,"YlGnBu")) # colors
  )

clu_method <- "ward.D2"

# The Heatmaps themselves
plei_cor_hm <- Heatmap(
  name="Spearman",
  cor(
    t(plei_tfs_zsco),
    method="sp" # Spearman correlation
  ),
  clustering_method_columns = clu_method,
  clustering_method_rows = clu_method,
  col=plei_tfs_heat_col,
  show_row_names = FALSE,
  show_column_names = FALSE
  )

plei_expr_zsco_hm <- Heatmap(
  name="norm\nexpr",
  t(scale(t(plei_tfs_cpm[,1:49])))+2, # +2 for visualisation purposes
  col=col_plei_expr_zsco_hm,
  show_row_names = FALSE,
  show_column_names = FALSE,
  cluster_rows=T,
  clustering_method_columns = "single",
  clustering_method_rows = "ward.D2",
  cluster_columns=F,
  top_annotation=clu_ha
  )
```

```{r echo = FALSE}
draw(plei_cor_hm)
draw(plei_expr_zsco_hm)
```

## Analysing TFs at the TF class level

We can explore the variability of TFs based at a TF class level.

```{r}
# boxplot CV per TF class

plei_tfs_class_cv <- merge(
  data.frame(
    id = plei_tfs$id,
    class = factor(
      plei_tfs$class,
      levels = unique(plei_tfs$class)
      )
    ),
  data.frame(
    id = rownames(plei_tfs_cpm),
    cv = apply(
      plei_tfs_cpm[,1:49],
      1,
      function(x){
        sd(x)/mean(x)
      }
    )
  ),
  by = 1
)[,2:3]

head(plei_tfs_class_cv)

```

```{r echo = FALSE}
tfs_cv_scatter <- 
  ggplot(plei_tfs_class_cv %>% group_by(class) %>%
           mutate( med = median(cv) ) %>%
           mutate( numgenes = log( table( as.character(class) ) ) ) ,
         aes(x = numgenes, y = med, label=class)) +
  geom_text(hjust=0, vjust=0, aes(colour = class))+
  scale_color_manual(
    values = rainbow(length(levels(plei_tfs_class_cv$class)))
    )+
  theme_classic()

tfs_cv_boxplot <- 
  ggplot(plei_tfs_class_cv %>% group_by(class) %>%
           mutate(med = median(cv)) %>%
           mutate(numgenes = log(table(as.character(class)))),
         aes(x = reorder(class, cv, FUN=median), y = cv)) +
  geom_boxplot(aes(fill = numgenes))+
  scale_fill_gradient(low="#00204D", high="#FFEA46")+
  theme_classic()+
  theme(
    axis.text.x = element_text(
      colour = rainbow(length(levels(plei_tfs_class_cv$class))),
      angle = 90
      )
    )
```

```{r echo = FALSE}
tfs_cv_scatter

pdf(
  file = "outputs/plots/plei_numtfs_mediancv_scatter.pdf",
  width = 8,
  height = 8
)
tfs_cv_scatter
dev.off()
```

```{r echo = FALSE}
tfs_cv_boxplot

pdf(
  file = "outputs/plots/plei_numtfs_cv_boxplot.pdf",
  width = 12,
  height = 7
)
tfs_cv_boxplot
dev.off()
```

With this we observe that there are a number of TF classes with very few assigned
genes in Pristina.

```{r echo = FALSE}
barplot(
  table(plei_tfs$class),
  col = rainbow(length(unique(sort(plei_tfs$class)))), # give them proper coloring like above
  las = 2,
  cex.names = 0.5,
  ylim = c(0,200)
  )

pdf(
  file = "outputs/plots/plei_numtfs_barplot.pdf",
  width = 12,
  height = 7
)
barplot(
  table(plei_tfs$class),
  col = rainbow(length(unique(sort(plei_tfs$class)))), # give them proper coloring like above
  las = 2,
  cex.names = 0.5,
  ylim = c(0,200)
  )
dev.off()

```

We can focus on those with a minimum of four or five genes.

```{r}
plei_tfs_mainclasses <- 
  names(
    table(plei_tfs$class)[
      table(plei_tfs$class) >= 4
      ]
    )
plei_tfs_mainclasses <- sort(plei_tfs_mainclasses) # sort alphabetically
plei_tfs_mainclasses
```

## ANOVA analysis of transcription factors

To find differences in TF class composition between cell clusters, we can resort to multivariate analysis two-way ANOVA to detect differences of TF expression between cell clusters, TF class, and the interaction of the two.

We will aggregate expression counts at the broad cell type level for this analysis.

```{r}
broadtypes_sort <- list( # using here the one from the excel
  piwi = c(1,2,3),
  epidermis = c(4,5,6,7,8,9,10,11),
  gut = c(12,13,14,15,16,17,18,19,20),
  muscle = c(21,22,23,24),
  neuron = c(25,26,27,28,29,30,31),
  blood = c(32,33),
  polycystin = c(34,35),
  eleo = c(36,37,38),
  chaetal = c(39,40),
  lipox = 41,
  vigil = 42,
  lumbrok = 43,
  carb = 44,
  secretory = c(45,46),
  arg = 47,
  ldlrr = 48,
  metaneph = 49
)
plei_cpm_broad <- sapply(
  X = broadtypes_sort,
  function(x) {if(length(x) > 1){rowSums(plei_cpm[,x])} else {plei_cpm[,x]}} #maybe should be done with counts_norm?
)

```

We generate our dataset for ANOVA analysis


```{r}
anova_data <- merge(
  plei_cpm_broad,
  plei_tfs,
  by.x = 0, by.y = 1)[,c(19,2:18)]

anova_data <- anova_data[anova_data$class %in% plei_tfs_mainclasses,]

anova_data <- reshape2::melt(anova_data)
colnames(anova_data) <- c("class","ctype","counts")
anova_data$class <- factor(anova_data$class, levels = sort(unique(anova_data$class)))
anova_data$ctype <- 
  factor(
    anova_data$ctype,
    levels = colnames(plei_cpm_broad)
    )

anova_groupby <- group_by(anova_data, class, ctype) %>%
  summarise(
    count = n(),
    mean = mean(counts, na.rm = TRUE),
    sd = sd(counts, na.rm = TRUE)
  )
```

A quick check of the looks of our dataset:

```{r}
summary(anova_data)
```

```{r}
summary(anova_groupby)
```

The anova analysis here

```{r}
# Two-way ANOVA proper
res.aov2 <- aov(counts ~ ctype + class + ctype:class, data = anova_data)
summary(res.aov2)
```

As we see there is significant interactions between TF class and cell cluster.

And an two-way interaction plot of the means of counts by class AND cell cluster.

```{r}
library(effects)

interaction.plot(
  x.factor = anova_data$ctype, trace.factor = anova_data$class, 
  response = anova_data$counts, fun = mean, 
  type = "b", legend = TRUE,
  pch=c(1,19),
  col = rainbow(35)
  )

plot(
  allEffects(res.aov2),
  multiline=TRUE
  )

```

Tukey test to compare the means of TF counts by class, by cell cluster

```{r}
# Tukey comparison of means
anova_comparisons_interaction <- TukeyHSD(res.aov2, which = "ctype:class")
anova_comparisons_interaction$`ctype:class` <- 
  anova_comparisons_interaction$`ctype:class`[
    anova_comparisons_interaction$`ctype:class`[,4] < 0.05 ,
    ]
```

From the interaction test, we extract the significant results for significant differences between the same TF class across different cell clusters.

```{r}
# put this all together with dplyr ...
# also think this better and implement a FUNCTION
anova_comparisons_interaction_DF <- data.frame(
  comparison = rownames(anova_comparisons_interaction$`ctype:class`),
  as.data.frame(
  anova_comparisons_interaction$`ctype:class`
  )
)
anova_comparisons_interaction_DF <- data.frame(
  anova_comparisons_interaction_DF,
  A = sub("-..*", "", anova_comparisons_interaction_DF$comparison),
  B = sub("..*-", "", anova_comparisons_interaction_DF$comparison)
)

anova_comparisons_interaction_DF$class_A <- 
  sub(".*:","",anova_comparisons_interaction_DF$A)
anova_comparisons_interaction_DF$class_B <- 
  sub(".*:","",anova_comparisons_interaction_DF$B)

anova_comparisons_interaction_DF <- 
  anova_comparisons_interaction_DF[
  anova_comparisons_interaction_DF$class_A == 
    anova_comparisons_interaction_DF$class_B ,
  ]

table(anova_comparisons_interaction_DF$class_A) # or class_B, its the same anyway
plei_tfs_signif_class <- names(table(anova_comparisons_interaction_DF$class_A))
```

A quick visualisation of what are the TF classes that explain the most differences between cell clusters:

```{r}
barplot(
  sort(table(anova_comparisons_interaction_DF$class_A)), # or class_B, its the same anyway
  col = rev(brewer.pal(9,"Spectral")),
  las = 2
  )

pdf(
  file = "outputs/plots/plei_classes_anovatukey.pdf",
  width = 8,
  height = 6
)
barplot(
  sort(table(anova_comparisons_interaction_DF$class_A)), # or class_B, its the same anyway
  col = rev(brewer.pal(9,"Spectral")),
  las = 2,
  main = "top classes explaining differences\nacross cell clusters"
  )
dev.off()
```

And we see that all these classes are already in the set of TF classes we will be exploring

```{r}
plei_tfs_signif_class %in% plei_tfs_mainclasses
```

## Visualising TF class prevalence across cell clusters

To visualise these differences in TF abundance at a TF class level, we retrieve the expression matrix at the class level. Below is a quick look too at the table (last columns, including class)

```{r}

plei_tfs_cpm_topclass <- 
  merge(
    plei_tfs_cpm,
    plei_tfs,
    by.x = 0,
    by.y = 1,
  ) %>% 
  column_to_rownames("Row.names") %>%
  filter(class %in% plei_tfs_mainclasses)

head(plei_tfs_cpm_topclass[,45:50])
```

We will generate a matrix to count how many genes of each TF class are expressed in each cell cluster.

```{r}
plei_tfs_ngenes <- 
  apply(
    plei_tfs_cpm_topclass[1:49],
    2,
    function(x){ ifelse(x > 0, 1, 0) }
    )
row.names(plei_tfs_ngenes) <- plei_tfs_cpm_topclass$class

plei_tfs_ngenes <-
  aggregate(
    plei_tfs_ngenes[, 1:49],
    by = list(Category = rownames(plei_tfs_ngenes)),
    FUN = sum) %>%
  arrange(Category) %>%
  column_to_rownames("Category")
```

We also generate a table of all the cpms per class and per cell cluster. We will aggregate (sum) all cpms by TF class, for each cell cluster separately. (We will also generate a normalised )

```{r}
plei_tfs_expgenes <-
  aggregate(
    plei_tfs_cpm_topclass[, 1:49],
    by = list(Category = plei_tfs_cpm_topclass$class),
    FUN = sum) %>%
  arrange(Category) %>%
  column_to_rownames("Category")
```

We divide the cpms/class by the numexp/class, thus retrieving the cpms per class normalised by the number of expressed genes from a given class.

```{r}
plei_tfs_expngenes <- 
  plei_tfs_expgenes / plei_tfs_ngenes
plei_tfs_expngenes <- apply(plei_tfs_expngenes, 2, function(x){ifelse(is.nan(x),0,x)})

plei_tf_EXPNGEN <- 
  apply(plei_tfs_expngenes, 2, function(x) {
    x / sum(x)
  })
```

A barplot to show the prominence/prevalence of TF classes on each cell cluster.

```{r echo = FALSE}
barplot_cluster(t(plei_tf_EXPNGEN)) # add color legend, make consistent with the rest of figures

pdf(
  file = "outputs/plots/plei_tfclass_barplotcluster.pdf",
  width = 14,
  height = 7
)
barplot_cluster(t(plei_tf_EXPNGEN))
dev.off()
```

## Saving the data

We will save the important bits for further analysis in the rest of markdowns.

```{r}
save(
  # cell type table(s)
  plei_counts,
  plei_ctypes,
  plei_ctypes_col,
  # gene expression data
  plei_cpm,
  plei_tfs_cpm,
  plei_tfs_cpm_topclass,
  # tf data
  plei_tfs,
  plei_tfs_mainclasses,
  plei_tfs_signif_class,
  # anova analysis
  res.aov2,
  anova_comparisons_interaction,
  anova_comparisons_interaction_DF,
  # visual annotations
  ctypes_rowAnno,
  clu_ha,
  #pick color palette for TFs
  # destination
  file = paste0(
    "./outputs/rda/",
    "02_tf_analysis.rda"
  )
)
```