---
title: "Iterating WGCNA graph network layouts"
author: "Alberto Perez-Posada"
date: "12/12/2022"
output: html_document
---

```{r setup, include=FALSE}
dir = "/mnt/sda/alberto/colabos/pristina/pristina_transcriptomic_landscape/"
fcha <- function(){ gsub("-","",Sys.Date()) }

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(eval = FALSE)
knitr::opts_knit$set(root.dir = dir)

options(scipen=999)
```

## About

This code can help deciding what thresholds can be chosen for further downstream analysis or plain visualisation of a graph (for a figure etc).
The printing layout code is based off of Katya Ognyanova's original igraph analysis tutorial on the website https://kateto.net/networks-r-igraph

## Load necessary packages and code:

We will load the igraph library and a custom-made function to parse the networks

## Load data

We first load our graph object.

```{r}
load("/path/to/wgcna/graph/igraph_object.rda")
```

## Declaring functions

We will first create a function to plot networks given a graph, an igraph layout object, an igraph layout name and a title (the latter two not needed but useful add the title in the plot).

Below is the code of the function just to document it in a bit more friendly way, but this function should be defined INSIDE the larger function (see below)

```{r plot_layout_function, eval = FALSE}
plot_layout <- function(graph,layout = NULL, l, title = NULL){
    plot(
      main = paste0(title, " ", layout),
      graph,
      vertex.size = 1.5,
      vertex.label = NA,
      vertex.color = "gray",
      edge.color = rgb(0,0,0,0.1),
      layout = l
    )
  }
```

The idea is to call this function from inside another function that parses a number of layouts and dumps the plot in a series of png files named after the layouts and a name of choice, in a path chosen by the user

```{r plot_the_network_function}
plot_the_network <- function(graph,layouts,savepng, graph_name, title_plot){
  
  plot_layout <- function(graph,layout, l, title){
    plot(
      main = paste0(title, " ", layout),
      graph,
      vertex.size = 1.5,
      vertex.label = NA,
      vertex.color = "gray",
      edge.color = rgb(0,0,0,0.1),
      layout = l
    )
  }
  
  for (layout in layouts) {
    
    print(layout) # Visual feedback
    
    l <- do.call(layout, list(graph)) 

    path_savepng <- paste0(
      savepng,
      fcha(),
      "_",
      layout,
      "_",
      graph_name,
      "_network.png"
      )
    
    png(
      path_savepng,
      wi = 1500, hei = 1500 , units = "px"
    )
    plot_layout(
      graph = graph,
      layout = layout,
      title = title_plot
      )
    dev.off()
  }
}
```

## Choosing layouts and thersholds

`igraph` offers a variety of layouts and one could even try to code their own layout algorithms in R, as long as the code returns a 2-col table with X and Y coordinates for the genes in the same order they are found inside the igraph object:

```{r layouts}
layouts <- c(
  # "layout_with_drl",
  # "layout_nicely",
  "layout_with_kk",
  "layout_components",
  "layout_with_fr",
  "layout_with_mds"
)
layouts
```

We create a number of thresholds to use for visualising the network.

```{r thresholds}
thresholds <- seq(0,0.95,by=0.05)
# thresholds <- seq(0.15,0.95, by = 0.005)
thresholds
```

## Iterating

We pick a saving directory and a potential name prefix for our files:

```{r}
save_dir = "./iterations_threshold/"
prefix = "plei_"
```

At the moment we **don't** have a list of attributes to add to the network.

```{r}
parsenetwork = FALSE
```

We can run the `for` loop below to iterate. For each threshold value, this for loop will call the `plot_the_network` function, to plot the subsetted network in different layouts, and will save the plots as png in our path of choice.

```{r forloop}
for (thresh in thresholds) {
  # Visual feedback
  print(thresh)
  
  # Create a graph name using the threshold value
  graphname <- paste0(prefix,thresh) 
  
  # Subset the graph based on the edges with a value above the threshold
  graph_iter <- subgraph.edges(
    plei_graph0,
    eids = which(E(plei_graph0)$weight >= thresh),
    delete.vertices = TRUE
    )
  
  if (parsenetwork == FALSE) {
    graph <- graph_iter
  } else {
    # We parse the network to add all the data needed to vertices and edges (not needed)
    parse <- ParseNetwork(graph_iter, attributes_list) # ATTR LIST TO BE PASSED DOWN TO THIS FUNCTION NEEDS TO BE ADDED TO THE MARKDOWN
  
    # We subset the parsenetwork object to retrieve the graph
    graph <- parse[[1]]
  }
  
  
  # More visual feedback: tell how many genes there are in the network
  print(
    paste0(graphname, ": ", vcount(parse[[1]]), " genes")
    )
  
  # We call the network plotting function
  plot_the_network(
    graph = graph,
    layouts = layouts,
    savepng = save_dir,
    graph_name = graphname,
    title_plot = graphname
    )
  }

```

The plots are saved in `save_dir` with the name `YYYYMMDD_{prefix}_{threshold}_{layout}_network.png`.

**Note** that this script can take a lot of time to run because it depends entirely on your graph object (size, number of edges etc.). Bear that in mind.
